---
title: "workshop3_metasm"
author: "CaterinaPalma"
date: "2026-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

# Workshop 3 

## Examine Reads 

# Step 1. Making Sub-Directory 
and then making a symbolic link. 

```bash 
mkdir raw_data
cd raw_data
ln -s /pvol/data/metagenome_assembly/reads_R10_0.5.fastq.gz raw_reads.fastq.gz
```

then back up out of raw data... 

```bash 
cd ..
``` 

### Check files are correct using "ls" 
This lists all files and directories in the project. Use the "-l" option to show extended details on each file which will allow you to see the source and target paths for the symbolic link. 

```bash
ls -lR .
``` 

### Read Length Distribution 
Taking shorter sequences and trying to connect them to longer sequences. 

```bash 
zcat raw_data/raw_reads.fastq.gz | head -n 40000 | bioawk -c fastx '{print $name,length($seq)}' > read_lengths.txt
``` 

### Creating a plot to visualise the read lengths: 

```{r}
library(tidyverse)

rl_data <- read_tsv("read_lengths.txt",col_names = c("id","length"), show_col_types = FALSE)

ggplot(rl_data, aes(x=length/1000)) + 
  geom_histogram(binwidth = 2) +
  xlab("Read Length Kb") + ylab("Number of Reads")
```
The above histogram Title: Figure 1: Histogram of read lengths in the mock nanopore sequencing dataset. 

### Calculating read N50 

The method we will use to calculate N50 is described by the following pseudo-code

Step 1: Rank all reads from longest to shortest
Step 2: Create a new column with the cumulative sum of read lengths (starting from longest read).
Step 3: Find the read at which the cumulative sum surpasses 50% of the total read volume (ie total sum of all read lengths)

```{r}
total_len <- sum(rl_data$length)

rl_data %>% 
  arrange(desc(length)) %>% 
  mutate(cumsum = cumsum(length)) %>% 
  filter(cumsum>total_len/2) %>% 
  head(n=1)
```

# Step 2. Assembly 
Assemble our reads! We will use flye for this task because it is probably the best assembler for long-read metagenome data.

1. Make flye directory 

```bash 
mkdir flye
cd flye
```

2. Making slurm 

save below into shell script    

#!/bin/bash   
#SBATCH --time=360   
#SBATCH --ntasks=8 --mem=4gb   
   
echo "Starting flye in $(pwd) at $(date)"
   
flye --nano-hq ../raw_data/raw_reads.fastq.gz --threads 8 --out-dir mock --meta
   
echo "Finished flye in $(pwd) at $(date)"

3. back to terminal... 

```bash 
sbatch run_flye.sh
```

My slurm number is "slurm-376" 

4. Monitor job output by... 

```bash 
tail -f slurm-376.out
``` 

Output: 

Total length:   18269961
Fragments:      41
Fragments N50:  1866219
Largest frg:    4681059
Scaffolds:      0
Mean coverage:  12

Once all finished. Go to flye then mock folder and export "assembly_graph.gfa" and put into Bandage. 

##### Question 3.1
What is the size and coverage of the largest contig in the assembly? 
- Largest size = 4,681,059 bp 
- Coverage (i.e. depth) = 21 

##### Question 3.2
Your assembly should contain at least one (probably several) large circular contigs. Based on size and coverage can you guess which contig corresponds to the genome of the most abundant species in our mock community?

- Largest contig corresponds to the Geitlerinema sp. PCC 7407 species. 

#### Chromosome lengths and GC content in the reference data 

Make a plain text file and save it as "gc.hwk" in the main folder with the below copy and pasted: 

# Bioawk script for calculating gc and printing name,length,size
BEGIN {
	OFS="\t"
}

{
	l=length($seq)
	ngc=gsub("[cgCG]","",$seq)
	nat=gsub("[taTA]","",$seq)
	print $name,l,ngc/(ngc+nat)
}

#### All the reference genomes for our mock community can be found at the path /pvol/data/mg/refs/. Take a look with ls

```bash 
ls /pvol/data/mg/refs/*.fna
``` 

Now test awk program by using it to print the GC content and contig lengths for all chromosomes in these files. Here we use cat to concatenate all the files together and then use a pipe to send their contents to our bioawk program

```bash 
cat /pvol/data/mg/refs/*.fna | bioawk -c fastx -f gc.awk
``` 

# Step 3. Basic Assembly QC
Previously we used gc.awk to print the length and GC content for reference sequences. Now we will do the same thing for our assembled contigs.

1. Make New Directory 

```bash 
mkdir assembly_check
cd assembly_check
``` 

2. Use your gc.awk script to print GC and lengths for the assembly

```bash 
cat ../flye/mock/assembly.fasta | bioawk -c fastx -f ../gc.awk 
``` 

3. Run the command again but this time send outputs to a file called assembly.gc.tsv.

```bash 
cat ../mock/assembly.fasta | bioawk -c fastx -f ../gc.awk > assembly.gc.tsv
```

We have GC content in "assembly.gc.tsv" and coverage depth in the "assembly_info.txt" file produced by flye. Let’s read those and join them together using R. 

```{r}
gc <- read_tsv("flye/assembly_check/assembly.gc.tsv",col_names = c("#seq_name","length","GC"))
info <- read_tsv("flye/mock/assembly_info.txt")
```

Then: 

```{r}
gc_info <- info %>% 
  left_join(gc)
```

Then: 

```{r}
gc_info %>% 
  ggplot(aes(x=GC)) + geom_histogram() + ylab("Number of contigs")
```
Figure 2: Histogram of GC content values values for assembled contigs in a flye assembly of the mock community dataset. 

A plot of GC vs coverage helps to identify clusters that have similar values for both these parameters. In making this plot we first filter the data to remove contigs with unrealistically high GC content and very high coverage depth (probably repeats). 

```{r}
gc_info %>% 
  filter(GC<0.8) %>% 
  filter(cov.<25) %>% 
  ggplot(aes(x=GC,y=cov.)) + geom_point(aes(size=length)) + ylab("Coverage depth")
```
Figure 3: Coverage and GC content of assembled contigs in the mock community assembly. Contig lengths are shown through point size.

Adding labels allows us to look at the contigs with highest coverage and compare these with the contigs in bandage. Notice how the contigs for Vibrio alfacsensis cluster together.

##### Question 3.4
In the plot of GC versus coverage there is a cluster of contigs with high GC content (>60%) and coverage depth around 7.5x. Which species are these likely to belong to?

```{r}
install.packages("ggrepel")
```


```{r}
library(ggrepel)
gc_info %>% 
  filter(GC<0.8) %>% 
  filter(cov.<25) %>% 
  ggplot(aes(x=GC,y=cov.)) + geom_point(aes(size=length)) + geom_label_repel(aes(label=`#seq_name`))
```
Figure 4: Coverage and GC content of assembled contigs in the mock community assembly. Contig lengths are shown through point size. Only contigs with more than 25x coverage and less than 80% GC coverage are shown.

# Assembly size vs reference size

Assessing out mock assembly because we should know every part of it, because it is a mock. 
We will do this from within our top-level project directory. Wherever you are you can change to this directory with the following command (assuming your project was called workshop3_metasm and you put it in your home directory). 

```bash 
cd flye
``` 

Firstly lets calculate the total length of our reference sequences. This command builds on one we ran earlier to list the size and GC content of reference chromosomes, but this time we have added another very simple awk program to take the sum of the second column.

```bash 
cat /pvol/data/mg/refs/*.fna | bioawk -c fastx -f gc.awk | awk '{sum+=$2}END{print sum}'
``` 

This soutput should give us a total length of the reference is around 24Mb...
My output: 24095709!!! 

# Step 4. Binning 

```bash 
mkdir metabat
cd metabat
```

Prepare input file 

```bash 
cat ../flye/mock/assembly_info.txt  | grep -v '#' | awk '{OFS="\t";print $1,$3}' > depths.txt
``` 

Next we need to create a new version of the assembly file in which the contigs have the same order as those in the depths.txt file. We do this using a samtools faidx which allows us to extract entries from a fasta file in whatever order we specify.

```bash 
cat depths.txt | awk '{print $1}' | xargs samtools faidx ../flye/mock/assembly.fasta > assembly_ordered.fasta
```

Finally, run metabat as follows: 

```bash 
metabat2 -i assembly_ordered.fasta -a depths.txt --cvExt -o bins/metabat --seed 5
```

Output: 
6 bins (13314935 bases in total) formed.

```bash 
grep '>' bins/*
``` 

# Step 5. Check bin quality with checkM 










